\documentclass{article}

\usepackage[preprint]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}

% ===== Architectural constants (change once, propagate everywhere) =====
\newcommand{\hdim}{512}
\newcommand{\nheads}{8}
\newcommand{\nlayers}{8}
\newcommand{\nctx}{4}
\newcommand{\ntokens}{64}
\newcommand{\vocabsize}{1{,}000}
\newcommand{\fsqlevels}{[8, 5, 5, 5]}
\newcommand{\gridsize}{8 \times 8}
\newcommand{\imgsize}{64 \times 64}
\newcommand{\fsqsigma}{0.9}
\newcommand{\fsqeps}{0.1}
\newcommand{\fsqdimweights}{[1.29, 0.85, 0.97, 0.89]}
\newcommand{\focalgamma}{2.0}
\newcommand{\nparams}{35\text{M}}

\title{Structured Label Smoothing over Finite Scalar Quantization\\for Real-Time World Models}

\author{%
  Florent Tariolle \\
  INSA Rouen Normandie \\
  \texttt{florent.tariolle@gmail.com} \\
}

\begin{document}

\maketitle

\begin{abstract}
World models that tokenize visual observations with Finite Scalar Quantization (FSQ) inherit a structured codebook where each code maps to a point on a discrete coordinate lattice.
Standard cross-entropy training ignores this structure, treating all incorrect predictions as equally wrong.
We introduce \emph{Structured Label Smoothing} (SLS), a training objective that distributes smoothing mass according to a Gaussian kernel over FSQ coordinate distances, weighted by per-dimension perceptual sensitivity.
We integrate SLS into a full Vision-Model-Controller pipeline for the platformer game Geometry Dash, combining an FSQ tokenizer, a causally-masked transformer with Adaptive Layer Normalization and 3D Rotary Position Embeddings, and a PPO controller trained entirely in dreamed rollouts.
The resulting system runs at 30\,FPS on consumer hardware.
% TODO: Add headline result once model is frozen (e.g., "achieving X\% progress on Level 1").
\end{abstract}

% ===========================================================================
\section{Introduction}
\label{sec:intro}
% ===========================================================================

Learning world models that predict future observations from actions is central to model-based reinforcement learning~\citep{ha2018world}.
Recent approaches tokenize visual frames into discrete codes and train autoregressive transformers to predict token sequences~\citep{micheli2023iris, hafner2023dreamerv3}.
Finite Scalar Quantization~\citep[FSQ;][]{mentzer2024fsq} offers a simple alternative to VQ-VAE codebooks: each latent vector is independently rounded to one of $L$ levels per dimension, yielding $\prod_d L_d$ implicit codes with no codebook collapse and no auxiliary losses.

A key property of FSQ, largely unexploited in downstream tasks, is that its codebook has a natural coordinate structure.
Each code maps to a point on a $D$-dimensional integer lattice, and the distance between two codes in this space is semantically meaningful: codes that differ by one step in one dimension represent visually similar patches.
Yet standard cross-entropy training treats all incorrect predictions as equally wrong, discarding this metric information entirely.

Label smoothing~\citep{szegedy2016rethinking} is a well-known regularizer that replaces hard targets with a mixture of the ground truth and a uniform distribution over classes.
We observe that for FSQ-tokenized world models, the uniform assumption is suboptimal: a token one lattice step away from the target is a near-miss, while a token at the opposite corner of the codebook is a gross error.
We propose \emph{Structured Label Smoothing} (SLS), which replaces the uniform component with a Gaussian kernel over FSQ coordinate distances, calibrated by per-dimension sensitivity weights derived from empirical patch reconstruction error.

We evaluate SLS within a complete Vision-Model-Controller (V-M-C) pipeline for Geometry Dash, a fast-paced platformer that demands frame-accurate predictions under a 33\,ms latency budget.
Our system consists of three components: (1)~an FSQ tokenizer that encodes $\imgsize$ Sobel edge maps into $\gridsize$ grids of discrete tokens, (2)~a $\nparams$-parameter causal transformer with Adaptive Layer Normalization (AdaLN-Zero) for action conditioning~\citep{peebles2023dit} and QK-norm for attention stability~\citep{esser2024sd3}, and (3)~a lightweight CNN controller trained via PPO entirely in dreamed rollouts~\citep{schulman2017ppo}.
The full pipeline achieves real-time inference at 30\,FPS on consumer hardware.

Our contributions are:
\begin{enumerate}
  \item \textbf{Structured Label Smoothing (SLS)}: a training objective that exploits FSQ coordinate geometry to distribute smoothing mass according to codebook distance, generalizable to any discrete latent space with a coordinate metric.
  \item \textbf{A real-time world model system}: a complete V-M-C pipeline for Geometry Dash running at 30\,FPS, combining FSQ tokenization, AdaLN-conditioned transformers, and PPO-in-dreams.
  % TODO: Add quantitative contribution once ablations are done.
\end{enumerate}

% ===========================================================================
\section{Related work}
\label{sec:related}
% ===========================================================================

\paragraph{Discrete world models.}
\citet{ha2018world} introduced the Vision-Model-Controller paradigm, encoding frames with a VAE, modeling dynamics with an RNN, and training a controller in dreamed rollouts.
IRIS~\citep{micheli2023iris} replaced the VAE with a discrete tokenizer and the RNN with an autoregressive transformer, showing that discrete tokens enable sample-efficient learning.
DreamerV3~\citep{hafner2023dreamerv3} demonstrated that discrete representations with symlog predictions generalize across diverse domains.
TWISTER~\citep{pignatelli2024twister} added action-conditioned contrastive predictive coding (AC-CPC) to improve temporal coherence.
Our work builds on these ideas but focuses on a previously unexploited axis: the structure of the discrete codebook itself.

\paragraph{Finite Scalar Quantization.}
FSQ~\citep{mentzer2024fsq} replaces learned codebooks with per-dimension rounding to fixed levels, eliminating codebook collapse and commitment losses.
Each code is a point on a $D$-dimensional integer lattice, giving the codebook a natural coordinate structure.
Prior work has used FSQ as a drop-in replacement for VQ-VAE but has not leveraged this coordinate structure during training.

\paragraph{Label smoothing.}
\citet{szegedy2016rethinking} proposed replacing one-hot targets with a uniform mixture to reduce overconfidence.
While effective as a regularizer, uniform smoothing ignores class relationships.
Knowledge distillation and soft-target methods~distribute mass based on teacher predictions, but require a separate teacher model.
We instead derive the target distribution from the codebook geometry itself, requiring no additional model.

\paragraph{Adaptive conditioning.}
FiLM~\citep{perez2018film} introduced adaptive affine conditioning via scale and shift parameters.
DiT~\citep{peebles2023dit} applied this as AdaLN-Zero with zero-initialized gates for stable training of diffusion transformers.
\citet{esser2024sd3} added QK-norm (RMSNorm on queries and keys per head) to stabilize attention logits in deep AdaLN models.
LeWorldModel~\citep{baranes2025leworldmodel} applied AdaLN to world model transformers for robot manipulation.
We adopt AdaLN-Zero with QK-norm for action conditioning, finding that QK-norm is essential for training stability (Section~\ref{sec:adaln}).

% ===========================================================================
\section{Method}
\label{sec:method}
% ===========================================================================

Our system follows the Vision-Model-Controller (V-M-C) paradigm~\citep{ha2018world}.
We describe each component below, with Structured Label Smoothing (Section~\ref{sec:sls}) as the primary contribution.

% ---------------------------------------------------------------------------
\subsection{FSQ tokenization (Vision)}
\label{sec:fsq}
% ---------------------------------------------------------------------------

We encode $\imgsize$ grayscale Sobel edge maps using an FSQ-VAE with levels $\fsqlevels$, producing $\vocabsize$ implicit codes.
The encoder applies three stride-2 convolutional blocks with residual connections, projecting to a 4-channel latent map of size $\gridsize$.
Each spatial position is independently quantized by applying $\tanh$, scaling to half-levels, and rounding, yielding $\ntokens$ tokens per frame.
The decoder mirrors the encoder with transposed convolutions.

Training minimizes MSE reconstruction loss with two regularizers from \citet{wang2020uniformity}: temporal slowness ($\lambda_\text{slow} = 0.1$) encouraging smooth latent trajectories across consecutive frames, and a uniformity loss ($\lambda_\text{unif} = 0.01$) preventing representational collapse.
The FSQ-VAE achieves full codebook utilization ($\vocabsize / \vocabsize$ active codes, perplexity 766).

% ---------------------------------------------------------------------------
\subsection{Structured Label Smoothing}
\label{sec:sls}
% ---------------------------------------------------------------------------

\paragraph{Motivation.}
Standard label smoothing~\citep{szegedy2016rethinking} replaces a one-hot target $\mathbf{e}_i$ with:
\begin{equation}
  q_\text{uniform}(j \mid i) = (1 - \varepsilon)\,\delta_{ij} + \frac{\varepsilon}{V - 1}(1 - \delta_{ij}),
  \label{eq:uniform_ls}
\end{equation}
where $V = \vocabsize$ is the vocabulary size and $\varepsilon$ controls the smoothing strength.
This distributes $\varepsilon$ uniformly across all alternatives, ignoring codebook geometry.

\paragraph{FSQ coordinate distance.}
An FSQ codebook with levels $\mathbf{L} = (L_1, \dots, L_D)$ assigns each code $i$ a coordinate vector $\mathbf{c}(i) = (c_1(i), \dots, c_D(i))$ on a $D$-dimensional integer lattice.
We define the weighted squared distance:
\begin{equation}
  d^2(i, j) = \sum_{d=1}^{D} \alpha_d \left( c_d(i) - c_d(j) \right)^2,
  \label{eq:fsq_dist}
\end{equation}
where $\alpha_d$ are per-dimension sensitivity weights.

\paragraph{Per-dimension calibration.}
To determine $\boldsymbol{\alpha}$, we measure how much a one-step perturbation in each FSQ dimension affects the decoded image.
For 800 randomly sampled patches, we perturb each dimension $d$ by $\pm 1$, decode both the original and perturbed codes, and compute the mean patch-wise MSE.
The weights $\alpha_d$ are set proportional to these empirical sensitivities, normalized to mean 1.
This yields $\boldsymbol{\alpha} = \fsqdimweights$ for our $\fsqlevels$ codebook, indicating that dimension~1 (with 8 levels) carries the most perceptual information.

\paragraph{Gaussian kernel smoothing.}
We replace the uniform distribution in Equation~\ref{eq:uniform_ls} with a Gaussian kernel over FSQ distances:
\begin{equation}
  w(i, j) = \exp\!\left( -\frac{d^2(i, j)}{2\sigma^2} \right), \quad j \neq i,
  \label{eq:gaussian_kernel}
\end{equation}
yielding the structured smooth target:
\begin{equation}
  q_\text{SLS}(j \mid i) =
  \begin{cases}
    1 - \varepsilon & \text{if } j = i, \\[4pt]
    \displaystyle \varepsilon \cdot \frac{w(i, j)}{\sum_{k \neq i} w(i, k)} & \text{otherwise}.
  \end{cases}
  \label{eq:sls}
\end{equation}
The full target matrix $\mathbf{Q} \in \mathbb{R}^{V \times V}$ is precomputed once and stored on GPU.
We use $\sigma = \fsqsigma$ and $\varepsilon = \fsqeps$.

\paragraph{Integration with focal loss.}
To handle class imbalance (background tokens dominate over obstacle and spike tokens), we combine SLS with focal loss~\citep{lin2017focal}:
\begin{equation}
  \mathcal{L}_\text{SLS-focal} = -\sum_{j=1}^{V} (1 - p_j)^\gamma \, q_\text{SLS}(j \mid i) \, \log p_j,
  \label{eq:sls_focal}
\end{equation}
where $p_j = \text{softmax}(\mathbf{z})_j$ and $\gamma = \focalgamma$.

% ---------------------------------------------------------------------------
\subsection{Transformer world model (Model)}
\label{sec:transformer}
% ---------------------------------------------------------------------------

The world model is a causal transformer that predicts the next frame's tokens given a context of $K = \nctx$ past frames and their associated actions.

\paragraph{Sequence construction.}
Each frame contributes $\ntokens$ visual tokens (from the $\gridsize$ FSQ grid) plus one status token (ALIVE or DEATH), for 65 tokens per frame.
A context window of $K + 1$ frames (including the target) produces a sequence of $325$ tokens.
All target-frame tokens are replaced with a learnable \texttt{[MASK]} embedding to prevent information leakage.

\paragraph{3D Rotary Position Embedding.}
We apply factored RoPE~\citep{su2024rope} encoding three positional axes: spatial row, spatial column, and temporal frame index.
The head dimension (64) is partitioned into three bands (22 row, 22 column, 20 time) with separate base frequencies ($\theta_\text{spatial} = 10$, $\theta_\text{time} = 10{,}000$).
Status tokens receive the grid center coordinates $(3.5, 3.5)$ to represent whole-frame summaries.

\paragraph{Block-causal attention.}
Within each frame block, attention is bidirectional (tokens attend to all positions in the same frame).
Across frames, attention is strictly causal: each frame attends only to current and past frames.

\paragraph{AdaLN-Zero action conditioning.}
\label{sec:adaln}
Rather than inserting action tokens into the sequence, we condition the transformer on the action history via Adaptive Layer Normalization~\citep{peebles2023dit, perez2018film}.
The $K$-step action sequence is one-hot encoded and projected through an MLP to produce per-layer scale, shift, and gate parameters that modulate each sublayer:
\begin{equation}
  \text{AdaLN}(\mathbf{x}) = (1 + \mathbf{s}) \odot \text{LayerNorm}(\mathbf{x}) + \mathbf{b}, \qquad
  \mathbf{y} = \mathbf{g} \odot f(\text{AdaLN}(\mathbf{x})),
  \label{eq:adaln}
\end{equation}
where $f$ is the attention or MLP sublayer and $(\mathbf{s}, \mathbf{b}, \mathbf{g})$ are predicted from the action context.
Gates are initialized with bias~$= 1$ to preserve residual flow at initialization, following our empirical finding that zero-initialized gates suppress both conditioning and learning (Section~\ref{sec:training_stability}).

\paragraph{QK-norm.}
We apply per-head RMSNorm to queries and keys before computing attention scores~\citep{esser2024sd3}.
This prevents attention logit overflow that otherwise causes training divergence in AdaLN-conditioned models, as we observed in our ablation of training stability (Section~\ref{sec:training_stability}).

\paragraph{AC-CPC auxiliary loss.}
Following \citet{pignatelli2024twister}, we add an Action-Conditioned Contrastive Predictive Coding loss.
Learned predictors map each frame's hidden representation and future actions to a shared embedding space, trained with InfoNCE~\citep{oord2018cpc} at temperature $\tau = 0.1$.

\paragraph{Output projection and total loss.}
The output projection shares weights with the input embedding (weight tying).
The total training loss is $\mathcal{L} = \mathcal{L}_\text{SLS-focal} + \lambda_\text{CPC} \mathcal{L}_\text{CPC}$ with $\lambda_\text{CPC} = 1.0$.

% ---------------------------------------------------------------------------
\subsection{PPO controller (Controller)}
\label{sec:controller}
% ---------------------------------------------------------------------------

The controller is a lightweight CNN that takes the predicted next-frame tokens and the transformer's hidden state as input, and outputs a jump probability and a state value estimate.

\paragraph{Architecture.}
Token IDs are embedded (16-dim), reshaped to a $16 \times \gridsize$ feature map, and processed by two convolutional layers with max-pooling, producing a 256-dim feature vector.
This is concatenated with the 256-dim transformer hidden state (projected from $\hdim$) and fed to separate actor and critic heads.
An auxiliary multi-token prediction head predicts the next 8 actions.
The full controller has approximately 40K parameters.

\paragraph{Training in dreams.}
The controller is first pretrained via behavioral cloning on expert demonstrations, then fine-tuned with PPO~\citep{schulman2017ppo} in dreamed rollouts.
At each iteration, 512 rollouts of up to 45 steps are generated by autoregressively sampling from the frozen world model.
Rollouts terminate when the predicted death probability exceeds 0.5.
Rewards are $+1$ per survival step with a $-0.2$ penalty per jump.
We use GAE~\citep{schulman2017ppo} with $\gamma = 0.995$, $\lambda = 0.95$, and clip ratio $\varepsilon_\text{PPO} = 0.2$.

% ===========================================================================
\section{Experimental setup}
\label{sec:setup}
% ===========================================================================

\paragraph{Environment.}
Geometry Dash is a rhythm-based platformer where the player controls a cube that moves at constant horizontal speed.
The only action is a binary jump.
Colliding with any obstacle causes instant death.
Levels are deterministic: the same action sequence always produces the same outcome, but frame-accurate timing is required.

\paragraph{Data collection.}
We collect 4,265 gameplay episodes (252K frames) including both intentional-death episodes (covering diverse failure modes) and 36 expert demonstrations (full level clears).
Vertical shift augmentation ($\pm 2, \pm 4$ pixels) expands the dataset 5$\times$.
Frames are preprocessed with Sobel edge detection and resized to $\imgsize$.

\paragraph{Training details.}
The FSQ-VAE is trained for 200 epochs with Adam (lr $= 10^{-3}$, batch size 2,048) on frame pairs.
The transformer is trained for 200 epochs with Adam (lr $= 4 \times 10^{-3}$, weight decay $0.01$, batch size 512) using cosine annealing with 5-epoch warmup.
We apply 5\% random token noise and 5\% FSQ-neighbor noise as data augmentation.
Death frames are oversampled $4\times$ via weighted sampling.
Mixed-precision training (FP16 with loss scaling on the transformer, BF16 on the FSQ-VAE) is used throughout.
All training runs on a single NVIDIA A100 GPU.

\paragraph{Evaluation metrics.}
% TODO: Fill in once metrics are finalized.
We evaluate on: (1)~next-token prediction accuracy and cross-entropy on held-out episodes, (2)~Contrastive Predictive Coding loss (CPC), (3)~Death prediction F1 score, and (4)~real-environment gameplay performance measured as percentage of level completed (mean over $N = 100$ automated runs).

% ===========================================================================
\section{Results}
\label{sec:results}
% ===========================================================================

% TODO: This entire section will be written after model freeze (deadline: May 31, 2026).
% Expected contents:
% - Table 1: Main results (level progress %, CPC, accuracy, Death F1) across model versions
% - Table 2: SLS ablation (with vs without, sigma sweep, dim weights vs uniform)
% - Table 3: Ablations (context length, action conditioning, death oversampling, MTP)
% - Table 4: Baselines (random, open-loop, heuristic, BC-only, frame-prediction-free)
% - Figure: dreamed rollout vs ground truth frames
% - Figure: inference latency breakdown per component
% - Table: per-GPU latency comparison

\textit{Results pending model freeze and ablation experiments.}

% ---------------------------------------------------------------------------
\subsection{Training stability of AdaLN variants}
\label{sec:training_stability}
% ---------------------------------------------------------------------------

% TODO: Turn this into a proper figure/table once all runs are final.

During development, we ran five AdaLN configurations to identify the minimal set of stabilization techniques required:

\begin{enumerate}
  \item \textbf{No gates} (4 params/layer): CPC reached 0.171 before diverging at epoch 44 due to attention logit explosion.
  \item \textbf{Zero-init gates} (6 params/layer): CPC plateaued at 0.276 with overfitting. Zero-initialized gates suppressed all residual flow, killing both conditioning and learning.
  \item \textbf{Gate bias $= 1$}: CPC reached 0.189, diverged at epoch 46. Restored learning speed but unbounded attention scores persisted.
  \item \textbf{Tanh bounding + gate bias $= 1$}: CPC reached 0.205, still diverged. Bounding modulation values to $[-1, 1]$ proved that the explosion source was not in the modulation parameters.
  \item \textbf{QK-norm + gate bias $= 1$}: Stable through 200 epochs, CPC $< 0.14$, surpassing all previous configurations. Per-head RMSNorm on queries and keys eliminated the attention logit overflow that was the true root cause.
\end{enumerate}

This progression demonstrates that in AdaLN-conditioned world model transformers, attention stability (via QK-norm) is more critical than modulation bounding.

% ===========================================================================
\section{Limitations}
\label{sec:limitations}
% ===========================================================================

Our evaluation is limited to a single game (Geometry Dash) with a binary action space and deterministic level layouts.
The generalization of SLS to environments with higher-dimensional actions, stochastic dynamics, or RGB observations remains to be validated.
The per-dimension calibration of $\boldsymbol{\alpha}$ requires a one-time empirical measurement that may not transfer across different FSQ configurations.
% TODO: Expand after results are in.

% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}
% ===========================================================================

% TODO: Write after results are finalized.
We introduced Structured Label Smoothing, a training objective that exploits the coordinate geometry of FSQ codebooks to improve next-token prediction in discrete world models.
Integrated into a full V-M-C pipeline with AdaLN-conditioned transformers and PPO-in-dreams, the system achieves real-time 30\,FPS gameplay on Geometry Dash.
SLS is applicable to any model that predicts over a discrete latent space with a coordinate metric, including image generation, video prediction, and planning.

\begin{ack}
  % TODO: Add compute acknowledgments (SLURM cluster, A100 access).
  % Geometry Dash is developed by RobTop Games.
\end{ack}

\bibliographystyle{plainnat}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

% TODO: Appendix sections to add:
% A. Full hyperparameter tables
% B. Additional dreamed rollout visualizations
% C. Per-position accuracy analysis
% D. Inference latency breakdown

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\input{checklist}

\end{document}
